
\documentclass[11pt]{exam}

\oddsidemargin=0.25truein \evensidemargin=0.25truein
\topmargin=-0.5truein \textwidth=6.0truein \textheight=8.75truein

%\RequirePackage{graphicx}
%\usepackage{hyperref}
\usepackage{comment}
%\usepackage{amssymb}
%\usepackage{amsfonts}
\usepackage{booktabs}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

% document starts here
\begin{document}
\parskip=\bigskipamount
\parindent=0.0in
\thispagestyle{empty}
{\large ECON-UB 233 \hfill Dave Backus @ NYU}

\bigskip\bigskip
\centerline{\Large \bf Quiz \#0:  Math Tools}
\centerline{Revised: \today}

{\it Please write your name in the upper right corner.
This quiz won't affect your grade, but 
it's a useful guide to me about your background and 
an accurate indication of the kinds of tools we'll be using.
Please write your answers in the space provided.
If you don't know an answer, perhaps you could suggest a way to get started.}

\begin{questions}
\question Consider the value of a consol,
a bond that pays a coupon $C$ at the end of every year, forever,
beginning next year.
If $q = 1/(1+r)$ is the discount factor (that's the term we'll use),
what is the present discounted value
\begin{eqnarray*}
    q C + q^2 C + q^3 C + \cdots \; ?
\end{eqnarray*}
What happens if $C=r$?

\vspace{2.5in}
\question Let consumption $c$ be positive and utility have the form
$u(c) = c^{1-\alpha}/(1-\alpha)$ for some parameter $\alpha > 0$.
What is the derivative of $u$ with respect to $c$?
The second derivative?
What are their signs?

\pagebreak
\question Consider a random variable $x$ that takes on the value one with probability $p$
and zero with probability $1-p$.
What is the expected value or mean of $x$?
The variance?  The standard deviation?


\vspace{3.5in}
\question What do the functions below represent?
\begin{eqnarray*}
    p(x) &=& (2 \pi \sigma^2)^{-1/2} \exp[ - (x-\mu)^2/2 \sigma^2] \\
    M(t) &=& \exp[ t \mu + t^2 \sigma^2/2] 
\end{eqnarray*}

\begin{comment}
\vspace{2in}
\question Suppose you have two matrices,
\begin{eqnarray*}
A \;\;=\;\; \left[
            \begin{array}{cc}
            1 & 2 \\ 2 & 1
            \end{array}
            \right] \;\;\mbox{ and }\;\;
B \;\;=\;\; \left[
            \begin{array}{ccc}
            1 & 2 & 3 \\ 4 & 5 & 6
            \end{array}
            \right] .
\end{eqnarray*}
What is AB?  BA?
\end{comment}
\end{questions}

\end{document}
