\documentclass[11pt]{exam}

\oddsidemargin=0.25truein \evensidemargin=0.25truein
\topmargin=-0.5truein \textwidth=6.0truein \textheight=8.75truein

%\RequirePackage{graphicx}
\usepackage[hypertex]{hyperref}
\usepackage{comment}
%\usepackage{amssymb}
%\usepackage{amsfonts}
\usepackage{booktabs}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

% document starts here
\begin{document}
\parskip=2\bigskipamount
\parindent=0.0in
\thispagestyle{empty}
{\large ECON-UB 233 \hfill Dave Backus @ NYU}

\bigskip\bigskip
\centerline{\Large \bf Quiz \#0:  Math Tools}
\centerline{(Started: July 19, 2011; Revised: \today)}

{\it Please write your name in the upper right corner.  
This quiz won't affect your grade, but it's an accurate indication of the kinds of tools
we'll be using and a useful guide to me about your background.  
Please write your answers in the space provided.
If you don't know an answer, perhaps you could suggest a way to get started.}

\begin{questions}
\question Consider the value of a consol,
a bond that pays a coupon $C$ at the end of every year, forever,
beginning next year.
If $q = 1/(1+r)$ is the discount factor (that's the term we'll use),
what is the present discounted value
\begin{eqnarray*}
    q C + q^2 C + q^3 C + \cdots \; ?
\end{eqnarray*}
What happens if $C=r$?

\vspace{2.5in}
\question Let consumption $c$ be positive and utility
$u(c) = c^{1-\alpha}/(1-\alpha)$ for some parameter $\alpha \geq 0$.
What is the derivative of $u$?
The second derivative?
What are their signs?

\pagebreak
\question Consider a random variable $x$ that takes on the value one with probability $p$
and zero with probability $1-p$.
What is the expected value or mean of $x$?
The variance?  The standard deviation?


\vspace{2in}
\question What does the function below represent?  
\begin{eqnarray*}
    p(x) &=& (2 \pi \sigma^2)^{-1/2} \exp[ - (x-\mu)^2/2 \sigma^2] 
\end{eqnarray*}

\vspace{2in}
\question Suppose you have two matrices,
\begin{eqnarray*}
A \;\;=\;\; \left[
            \begin{array}{cc}
            1 & 2 \\ 2 & 1
            \end{array}
            \right] \;\;\mbox{ and }\;\;
B \;\;=\;\; \left[
            \begin{array}{ccc}
            1 & 2 & 3 \\ 4 & 5 & 6
            \end{array}
            \right] .
\end{eqnarray*}
What is AB?  BA?
\end{questions}

\end{document}
