\documentclass[11pt]{exam}

\oddsidemargin=0.25truein \evensidemargin=0.25truein
\topmargin=-0.5truein \textwidth=6.0truein \textheight=8.75truein

%\RequirePackage{graphicx}
\usepackage{comment}
\usepackage{verbatim}
\usepackage{booktabs}
\usepackage{hyperref}
\urlstyle{rm}   % change fonts for url's (from Chad Jones)
\hypersetup{
    colorlinks=true,        % kills boxes
    allcolors=blue,
    pdfsubject={ECON-UB233, Macroeconomic foundations for asset pricing},
    pdfauthor={Dave Backus @ NYU},
    pdfstartview={FitH},
    pdfpagemode={UseNone},
%    pdfnewwindow=true,      % links in new window
%    linkcolor=blue,         % color of internal links
%    citecolor=blue,         % color of links to bibliography
%    filecolor=blue,         % color of file links
%    urlcolor=blue           % color of external links
% see:  http://www.tug.org/applications/hyperref/manual.html
}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\newcommand{\var}{\mbox{\it Var\/}}

%\printanswers

% document starts here
\begin{document}
\parskip=\bigskipamount
\parindent=0.0in
\thispagestyle{empty}
{\large ECON-UB 233 \hfill Dave Backus @ NYU}

\bigskip\bigskip
\centerline{\Large \bf Lab Report \#2: Certainty Equivalents \& Portfolio Choice}
\centerline{(Started: July 19, 2011; Revised: \today)}

\bigskip
{\it Due at the start of class.
You may speak to others, but whatever you hand in should be your own work.}

\begin{questions}

\begin{solution}
Most of the answers are computed in the Matlab
program listed at the end and posted
\href{http://pages.stern.nyu.edu/~dbackus/233/hw2_s12.m}{here}.
\end{solution}

%-----------------------------------------------------------------------
\question (lognormal risks)
Our mission is to compute the {\it risk penalty\/}
associated with lognormal risks and power utility.
What risks matter?  What role does the risk aversion parameter play?

To be concrete, let $x = \log c \sim \mathcal{N}(\kappa_1, \kappa_2)$
and $u(c) = c^{1-\alpha}/(1-\alpha)$ for $\alpha> 0$.
The risk penalty is
\begin{eqnarray*}
    \mbox{\it rp\/} &=& \log (\bar{c}/\mu)
                \;\;=\;\; \log \bar{c} - \log \mu ,
\end{eqnarray*}
where $\bar{c} = E (c) = E( e^{x})$ and $\mu$ is the certainty equivalent.
We'll compute the terms one by one.

\begin{parts}
\part What is the cumulant generating function (cgf) of $x = \log c$?
\part Use the cgf to compute $\bar{c}$.
\part Use the cgf to compute the certainty equivalent $\mu$.
\part What is the risk penalty?  How does it depend on risk?  Risk aversion?
\part Suppose $\kappa_2 = 0.02^2$.  What is the risk penalty if $\alpha = 2$?
$\alpha = 10$?  $\alpha = 20$?
\end{parts}

\begin{solution}
Here risk is the variance $\kappa_2$ and risk aversion is the preference
parameter $\alpha$.
The risk penalty is $\alpha\kappa_2/2$, so both risk and risk aversion matter.
The cgf is used to compute all of the components:
$\bar{c}$, $\mu$, and $\mbox{\it rp\/}$.
See the Matlab program for details.
\end{solution}

%-----------------------------------------------------------------------
\question (exponential risks)
Now consider the same problem when $x = - \log c $ is exponential,
a tractable nonnormal distribution.
The connection between $c$ and $x$ implies $c = e^{\log c} = e^{-x}$.
The pdf is
\begin{eqnarray*}
    p(x)  &=& \lambda e^{-\lambda x}
\end{eqnarray*}
for $x \geq 0$ and $\lambda > 0$.
The cgf is
\begin{eqnarray*}
    k(s)  &=& - \log \left( 1 - s/\lambda \right) .
\end{eqnarray*}
In terms of $x$, the mean of $c$ is $\bar{c} = E(c) = E (e^{-x})$.

\begin{parts}
\part Use the cgf to compute the variance, skewness, and excess kurtosis
of $x$.  How do they compare to those of the normal distribution?
\part Use the cgf to compute $\bar{c}$.
\part Choose the value of $\lambda$ that sets
the variance equal to $0.02^2$, the number we used earlier.
\part
What is the risk penalty if $\alpha = 2$? $\alpha = 10$?
$\alpha = 20$?
How does your answer differ from the previous one?  Why?
\end{parts}

\begin{solution}
Same approach as the previous problem.
\begin{parts}
\part The first four cumulants are
$\kappa_1 = 1/\lambda$,
$\kappa_2 = 1/\lambda^2$,
$\kappa_3 = 2/\lambda^3$,
and $\kappa_4 = 6/\lambda^4$.
Skewness and excess kurtosis are therefore
$\gamma_1 = 2$ and $\gamma_2 6$.

\part If $k(s)$ is the cgf of $x$,
then $\log \bar{c} = k(-1)$ ($k$ evaluated at $s=-1$):
$k(-1) = - \log (1 + 1/\lambda)$.

\part We solve $0.02^2 = 1/\lambda^2$ or $\lambda = 1/0.02 = 50$.


\part The risk penalties are
%
\begin{center}
\begin{tabular}{lccc}
        & $\alpha=2$  & $\alpha=10$  & $\alpha=20$ \\
\midrule
lognormal    &  0.0004 & 0.0020 & 0.0040 \\
exponential  &  0.0004 & 0.0022 & 0.0054
\end{tabular}
\end{center}
%
The point is that departures from normality matter, although they're not
especially large here.
The numbers suggest that the impact increases with risk aversion $\alpha$.
We could elaborate on this using the power series expansion of the cgf,
which would show you that the $j$th cumulant is multiplied
by $(-\alpha)^j/j!$, so as we increase $\alpha$ some of these terms can increase
dramatically.
Amir Yaron (Wharton prof) refers to this as a ``bazooka,''
since it allows you to take a little nonnormality
and generate enormous risk premiums.
More on this kind of thing later in the course.
\end{parts}
\end{solution}


%-----------------------------------------------------------------------
\question (2-state portfolio choice)
Portfolio choice problems are notoriously unfriendly,
but we can get a sense of their properties with an example.
The agent's consumption-saving-portfolio-choice problem is
\begin{eqnarray*}
   \max_{c_0, a} &&  u(c_0) + \beta \sum_z p(z) u[c_1(z)] \\
        s.t. &&  c_1(z)\;\;=\;\; (y_0-c_0)[(1-a) r^1 + a r^e(z)] .
\end{eqnarray*}
The idea is that we have two assets,
a riskfree asset with gross return $r^1$ (``$1$'' for one-period bond)
and a risky asset with gross return $r^e$  (``$e$'' for equity).
We invest a fraction $a$ of our saving in equity and the complementary
fraction $1-a$ in the riskfree bond.
If $a>1$, the agent has a levered position.
Note that this is slightly different from the version we used in class.

For most of this question, we'll deal with a special case:
two states, $z=1$ and $z=2$;
equally likely, $p(1) = p(2) = 1/2$;
and power utility, $u(c) = c^{1-\alpha}/(1-\alpha)$.
Parameter values include
$\beta = 1/1.1$,
$r^1 = 1.1$,
$r^e(1) = 1.0$, $r^e(2) = 1.4$,
$ y_0 = 1$.

\begin{parts}
\part What are the mean and variance of the return on equity?
\part What are the implied prices of Arrow securities?
Hint:  Recall that Arrow securities pay off in one state only.
Our assets are combinations of Arrow securities,
so it's a question of unbundling them.
\part What are the first-order conditions for $c_0$ and $a$?
Show that with power utility,
the latter can be determined without knowing $c_0$.
\part Solve these conditions for $\alpha = 5$.
What values do you get for $a$, $c_0$, $c_1(1)$, and $c_1(2)$?
Comment:  You can do this by hand, but I did it numerically
by varying $a$ until its first-order condition was satisfied.
\part How does your answer change if you use $\alpha = 2$?
Does the difference make sense to you?
\end{parts}

\begin{solution}
\begin{parts}
\part The mean is 1.2 (the average of 1.0 and 1.4)
and the variance is $0.2^2 = 0.04$.

\part
Let $q^A(z)$ be the price of the Arrow security that pays one
in state $z$.
A one-period bond pays one in each state, so its price is the sum:
\begin{eqnarray*}
    q^1 \; (=1/r^1)  &=&  q^A(1) + q^A(2)  .
\end{eqnarray*}
Equity is a little more complicated, because we have to agree on
units.  Let's choose units to that the price of a share is one
unit of date-0 consumption.
Then it pays off 1.0 units of date-1 consumption in state $z=1$
and 1.4 units in state $z=2$:
\begin{eqnarray*}
    q^e \; (=1)  &=&  1.0 q^A(1) + 1.4 q^A(2)  .
\end{eqnarray*}
Now it's a matter of algebra to find
$q^A(1) = 0.6818$ and $q^A(2) = 0.2273$.


\part The first-order conditions are
\begin{eqnarray*}
    c_0:  &&  u'(c_0) \;\;=\;\; \beta \sum_z p(z)
%            u' \left\{  (y_0-c_0) [ (1-a)r^1 + a r^e(z)] \right\}
            u'[c_1(z)] [ (1-a)r^1 + a r^e(z)] \\
    a:    && 0 \;\;=\;\; \beta \sum_z p(z) u'[c_1(z)] [r^e(z) - r^1 ]  .
\end{eqnarray*}
Written out more completely, the latter becomes
\begin{eqnarray*}
    0 &=& \beta \sum_z p(z) (y_0-c_0)^{1-\alpha} [ (1-a)r^1 + a r^e(z)]^{-\alpha} [r^e(z) - r^1 ] \\ &=&  \sum_z p(z) [ (1-a)r^1 + a r^e(z)]^{-\alpha} [r^e(z) - r^1 ] .
\end{eqnarray*}
which depends on $a$ but not $c_0$.

\part I did this numerically.
We'll talk about numerical methods for solving equations shortly (``root-finding''),
but the simplest way is to compute the right side of the previous equation
for a grid of values for $a$.
We take the value that produces the first-order condition closest to zero.
A grid of 0.01 gives us $a=1.70$ when $\alpha=2$.

Now that we know $a$, we can find $c_0$ from the first first-order condition:
\begin{eqnarray*}
    c_0^{-\alpha}  &=&  \beta (y_0-c_0)^{-\alpha}
            \sum_z p(z) [ (1-a)r^1 + a r^e(z)]^{1-\alpha}  .
\end{eqnarray*}
We compute the sum on the rhs first --- denote it by $S$ (for sum).
Then we solve for $c_0$:
\begin{eqnarray*}
    (y_0-c_0)/c_0 &=& (\beta S)^{-1/\alpha}
        \;\;\Rightarrow\;\;
        c_0  \;\;=\;\; y_0 /[ 1 + (\beta S)^{-1/\alpha}] .
\end{eqnarray*}
Second-period consumption follows from the budget constraint:
$c_1(z)=  (y_0-c_0)[(1-a) r^1 + a r^e(z)] $.

The numbers are listed below:
\begin{center}
\begin{tabular}{lccccc}
                &  $a$  &  $S$    & $c_0$   &  $c_1(1)$ &  $c_1(2)$  \\
\midrule
$\alpha = 2$    & 1.702 & 0.8482  & 0.4676  & 0.4951    & 0.8576 \\
$\alpha = 5$    & 0.637 & 0.6135  & 0.4708  & 0.5484    & 0.6832
\end{tabular}
\end{center}


\part When $\alpha=5$, we get $a=0.64$ by the same method.
The punchline:  when we increase risk aversion, we hold less of the risky asset
and less risky second-period consumption as a result.

\end{parts}
\end{solution}
\end{questions}

\vfill \centerline{\it \copyright \ \number\year \
NYU Stern School of Business}

\end{document}


\pagebreak
Matlab program:
\verbatiminput{../Matlab/hw2_s12.m}

\end{document}



